---
title: 'Graph-Level Indices and CUG Tests'
author: Divya Krishnan
date: 'Wednesday, April 22nd, 2015'
output: pdf document
---

\

### Graph-Level Indices and CUG Tests[^1] ###

[^1]: Problems originally written by C.T. Butts (2009).
\  
<hr>

```{r}
# Adding libraries
library(statnet)
library(sna)
library(network)
library(knitr)
# Setting working directory
#setwd("C:/Users/Divya/Desktop/R programming/Problem sets")
# Loading relevant datasets 
load("sampson.Rdata")
load("Cow.Rdata")
load("Graph_level_indices.Rdata")
# Listing variables 
ls()
```

\  

#### Graph-Level Indices ####

Consider the Sampson monk data[^2]. Sampson collected various relationships between several monks at a monastery. Suppose we divide the types of social ties into positive and negative relationship types as follows:

* Positive Relationships: Esteem, Influence, LikeT1, LikeT2, LikeT3, and Praise
* Negative Relationships: Disesteem, NegInfluence, Dislike, and Blame

Using a vector permutation test, evaluate the questions below.

[^2]: F. S. Sampson. A novitiate in a period of change: An experimental and case study of social relationships. PhD thesis, Cornell University. 1968.


##### (a) ##### 

Are positive relations more reciprocal (relative to density) than negative ones? 
\
Answer: There is a statistically significant correlation between reciprocity (relative to density) and whether the network is positive or negative. Hence, we can say that positive relations are more reciprocal than negative ones. 

```{r}
perm.cor.test<-function(x,y,niter=5000){  #Define a simple test function
  c.obs <- cor(x,y,use="complete.obs")
  c.rep <- vector()
  for(i in 1:niter)
    c.rep[i] <- cor(x,sample(y),use="complete.obs")
  cat("Vector Permutation Test:\n\tObserved correlation: ",c.obs,"\tReplicate quantiles (niter=",niter,")\n",sep="")
  cat("\t\tPr(|rho|>=|obs|):",mean(abs(c.rep)>=abs(c.obs)),"\n")
  invisible(list(obs=c.obs,rep=c.rep))
  return(mean(abs(c.rep)>=abs(c.obs)))
}

# Printing the sampson network to understand how the different networks are arranged
# sampson
# Vector for marking the positive and negative networks
pos.neg<-c(1,0,1,0,1,1,1,0,1,0)
# Calculating the reciprocity relative to density for each network within the sampson dataset
sam.recip<-grecip(sampson,measure="edgewise.lrr")
# Printing the reciprocities of each network
sam.recip
# Permutation test to check if the network's attribute of being positive or negative has an affect on the reciprocity of the network
perm.cor.test(sam.recip,pos.neg)

```

##### (b) #####

Are positive relations more transitive (relative to density) than negative ones?
\
Answer: There is a statistically significant correlation between transitivity (relative to density) and whether the network is positive or negative. Hence, we can say that positive relations are more transitive than negative ones. 

```{r}

# Calculating the transitivity relative to density for each network within the sampson dataset
sam.trans<-log(gtrans(sampson)/gden(sampson))
# Printing the transitivity of each network
sam.trans
# Permutation test to check if the network's attribute of being positive or negative has an affect on the transitivity of the network
perm.cor.test(sam.trans,pos.neg)

```

##### (c)  #####

Discuss the findings from part (a) and part (b).
\
Answer: The p value obtained from the permutation test of reciprocity and transitivity with respect to positive/negative network attribute, is close to zero (approximately 0.05). This means that the correlation between reciprocity and positive/negative attribute of the network and correlation between transitivity and positive/negative attribute of the network is statistically significant. Hence, the fact that the network is a positive one, affects its reciprocity and transitivity. 
\
Another way of explaining the above statement is that positive networks are more reciprocal, that means, if monk1 likes monk2 then there are higher chances of monk2 liking monk1 back. Since this is edgewise reciprocity, it indicates that there are less chances of monk1 liking monk2 but monk2 not liking monk1 back. Hence, in most cases the monks like each other or they have no relatonship with respect to liking. This is not the case for negative networks. Negative networks would tend to be less reciprocal, that means, if monk1 dislikes monk2, there are less chances of monk2 disliking monk1.
\
The above statement also states that positive networks are more transitive, that means, if monk1 likes monk2 and monk2 likes monk3, then there are chances of monk1 liking monk3. This is not the case for negative networks. Negative networks would tend to be less transitive, that means, if monk1 dislikes monk2, and monk2 dislikes monk3, then there are less chances of monk1 disliking monk3.
\  
\
\
#### Random Graphs ####


##### (a)  Generating Random Graphs #####

Generate 100-node random directed graphs with expected densities of 0.0025, 0.005, 0.01, 0.015, 0.02, and 0.025, with at least 500 graphs per sample. Remember the `rgraph` function can draw more than one graph at a time. Plot the average Krackhardt connectedness, dyadic reciprocity, and edgewise reciprocity as a function of expected density. Use these to describe the baseline effect of increasing density on network structure.

```{r}
par(mfrow=c(1,1))
# Creating random graphs based on the given densities 
rand<-rgraph(100,500,c(0.0025, 0.005, 0.01, 0.015, 0.02, 0.025),mode="digraph")

# Plotting Krackhardt connectedness as a function of expected density
plot(gden(rand),connectedness(rand),main="Krackhardt connectedness",xlab="Density",ylab="Connectedness")

```

\
Answer: Effect of density on Krackhardt connectedness - From the plot on Krackhardt connectedness, we observe that as density slowly increases from zero, the connectedness increases at a very slow rate. This is because as we add edges, the density increases but the connectedness does not increase much as the network is still very disconnected. But as density increases beyond 0.005 till about 0.015, the connectedness increases steeply. At this time, the network is changing from a disconnected network to a connected network. Once the network is connected, then increasing the number of edges, increases density but may not increase connectedness that much, as the network is already connected. Hence we see that as density increases, Krackhardt connectedness increases, though not in a perfectly linear fashion.
\

```{r}
# Plotting dyadic reciprocity as a function of expected density
plot(gden(rand),grecip(rand,measure="dyadic"),main="Dyadic Reciprocity",xlab="Density",ylab="Dyadic Reciprocity")

```
\
Answer: Effect of density on Dyadic Reciprocity - From the plot on Dyadic Reciprocity, we observe that as density increases, dyadic reciprocity decreases. There is a strong negative correlation between density and dyadic reciprocity. When density is zero, the number of null dyads is maximum and hence making dyadic reciprocity very high. As density increases, the number of null dyads decreases. It is also possible that more asymetric dyads are increasing as density increases and hence the dyadic reciprocity decreases. But there is a possibility that as density increases to very high levels, the dyadic reciprocity will also increase as the number of mutual dyads will increase. Hence we estimate that if the density range is increased, we may see a U-shaped relationship between density and dyadic reciprocity.
\

```{r}
# Plotting edgewise reciprocity as a function of expected density
plot(gden(rand),grecip(rand,measure="edgewise"),main="Edgewise Reciprocity",xlab="Density",ylab="Edgewise Reciprocity")

```

\
Answer: Effect of density on Edgewise Reciprocity - The edgewise reciprocity when plotted against density is not a smooth function. Hence, we see the jumps in the plot. It is possible as density increases, the mutual dyads either increase or do not change. If the mutual dyads increase, the edgewise reciprocity will increase else it will not have much difference. 
\
\

##### (b) Comparing GLIs #####

Generate 500 random directed graphs whose dyad census is the same as that of `kaptail.ins`. Plot histograms for total degree centralization, betweenness centralization, transitivity, and Krackhardt connectedness from this random sample. On your plot mark the observed values of these statistics (from the `kaptail.ins` data) using a vertial line You mgith find the `abline` function helpful here. Try modifying the `lwd` argument to the plot function to make the vertical line stand out. How do the replicated graphs compare to the observed data.

```{r}
par(mfrow=c(1,1))
# Calculating the dyad census for kaptail.ins
dcensus<-dyad.census(kaptail.ins)
# Generating random graphs from N,m family by giving the exact dyad census
nm<-rguman(500,39,dcensus[1],dcensus[2],dcensus[3],method="exact")
# Generating the histogram for total degree centralization
hist(centralization(nm,degree,mode="digraph"),main="Total Degree Centralization",xlab="")
# Marking the observed values
abline(v=centralization(kaptail.ins,degree,mode="digraph"),col=2,lwd=3) 
# Generating the histogram for betweenness centralization
hist(centralization(nm,mode="digraph",betweenness),main="Betweenness Centralization",xlab="")
# Marking the observed values
abline(v=centralization(kaptail.ins,mode="digraph",betweenness),col=2,lwd=3) 
# Generating the histogram for transitivity
hist(gtrans(nm,mode="digraph"),main="Transitivity",xlab="")
# Marking the observed values
abline(v=gtrans(kaptail.ins,mode="digraph"),col=2,lwd=3) 
# Generating the histogram for Krackhardt connectedness
hist(connectedness(nm),main="Krackhardt Connectedness",xlab="")
# Marking the observed values
abline(v=connectedness(kaptail.ins),col=2,lwd=3) 

```

\
Answer:
\

Total Degree Centralization -  The observed value of Kaptail.ins is towards the upper tail of the plot, which means most of the random graphs of similar structure will lie to the left of Kaptail.ins in terms of total degree centralization. Hence, we can say that the total degree centralization observed in Kaptail.ins is more than what is usually observed in a random graph of N,m family.
\ 

Betweenness Centralization -  The observed value of Kaptail.ins lies in the middle of the plot. Hence, we can say that the betweenness centralization observed in Kaptail.ins is similar to what is usually observed in a random graph of N,m family.
\ 

Transitivity -  The observed value of Kaptail.ins is in the extreme upper end of the plot, which means almost all of the random graphs of similar structure will lie to the left of Kaptail.ins in terms of transitivity. Hence, we can say that the transitivity observed in Kaptail.ins is much more than what is usually observed in a random graph of N,m family. Since the transitivity observed in Kaptail.ins is so much higher than usually observed in a random graph, there is a possibility of investigation into the underlying conditions that led to so much transitivity in Kaptail.ins network.
\ 

Krackhardt Connectedness -  The observed value of Kaptail.ins is towards the lower tail of the plot, which means that most of the random graphs of similar structure will lie to the right of the Kaptail.ins in terms of Krackhardt's connectedness. Hence, we can say that the Krackhardt's connectedness observed in Kaptail.ins is less than what is usually observed in a random graph of N,m family. 
\
\
\
#### Testing Structural Hypotheses ####

Consider the following set of propositions, which may or may not be true of given dataset. For each, do the following:

1. Identify a statistic (e.g. GLI) whose value should deviate from a random baseline if the proposition is true. 
2. Identify the approporate baseline distribution to which the statistic should be compared. 
3. Determine whether the proposition implies that the statistic should be greater or lower than its baseline distribution would indicate. 
4. Conduct a conditional uniform graph test based on your conclusions in 1-3. In reporting your results, include appropriate summary output from the `cug.test` function as well as the resulting distributional plots. Based on the results, indicate whether the data appears to support or undermine the proposition in question. Be sure to justify your conclusion. 

##### (a) #####

In militarized interstate disputes, hostile acts are disproportionately likely to be response to in kind. 

```{r}
# CUG test for edgewise reciprocity conditioned on edges
cug<-cug.test(mids_1993,grecip,cmode="edges",FUN.arg=list(measure="edgewise"))
cug
plot(cug)

```

* Statistic: Edgewise Reciprocity
* Baseline Distribution: Since the proposition considers hostile acts, hence we are Conditioning on density (which means we are conditioning on size and edges)
* Direction of Deviation: Higher than baseline distribution
* Conclusion and Discussion: Since we are looking at a one-tailed test, we consider the p value of Pr(X>=obs), which is 0. Since this value is less than 0.05(the value of alpha), we reject the null hypothesis. This means that the observed edgewise reciprocity of the mids_1993 network deviates from the baseline distribution. This makes the proposition true. Hence, in militarized interstate disputes, hostile acts are disproportionately likely to be response to in kind. We may like to consider investigating on the factors that could have led to this higher observed reciprocity in this network. 


##### (b) #####

When engaging in disputes, nations behave in accordance with the notion that "the enemy of my enemy is not my enemy".

```{r}
# CUG test for transitivity conditioned on edges
cug<-cug.test(mids_1993,gtrans,cmode="edges")
cug
plot(cug)
```

* Statistic: Transitivity
* Baseline Distribution: Since the proposition considers engaging in disputes, hence we are Conditioning on density (which means we are conditioning on size and edges)
* Direction of Deviation: Lower than baseline distribution
* Conclusion and Discussion: Since we are looking at a one-tailed test, we consider the p value of Pr(X<=obs), which is 0.959. Since this value is not less than 0.05(the value of alpha), we fail to reject the null hypothesis. This makes the above said proposition false. Hence, when engaging in disputes, nations do not behave in accordance with the notion that "the enemy of my enemy is not my enemy". This also means that the transitivity observed in mids_1993 network is consistent with the baseline distribution.

##### (c) #####

Given the number of disputes at any given time, a small number of nations will receive a disproportionate share of aggressive acts. 

```{r}
# CUG test for indegree centralization conditioned on edges
cug<-cug.test(mids_1993,centralization,cmode="edges",FUN.arg=list(FUN=degree,
cmode="indegree"))
cug
plot(cug)

```

* Statistic: Indegree centralization
* Baseline Distribution: Since the proposition fixes the number of disputes at a given time, hence we are Conditioning on density (which means we are conditioning on size and edges) 
* Direction of Deviation: Higher than baseline distribution
* Conclusion and Discussion: Since we are looking at a one-tailed test, we consider the p value of Pr(X>=obs), which is 0. Since this value is less than 0.05(the value of alpha), we reject the null hypothesis. This makes the above said proposition true. Hence, given the number of disputes at any given time, a small number of nations will receive a disproportionate share of aggressive acts. Since the observed indegree centralization deviates from the baseline distribution, we may want to investigate the underlying special conditions that led to the creation of the network.
